### 2014-01-31 manuscripts/dpp-msbayes/dpp-msbayes-main.tex
More tweaks to intro of DPP manuscript.

\section*{Introduction}
Understanding the processes that generate biodiversity and regulate
community assembly is a major goal of evolutionary biology.
Given the dynamic nature of our planet, a major component of this goal is
understanding how geological and climatic processes can affect the
evolutionary history of whole groups of co-distributed species and their
associated microbiota, including pathogens.
Such large-scale processes predict patterns of shared evolutionary history
across affected populations of organisms, e.g., temporally clustered
divergences across co-distributed species.
At recent timescales, such events should leave a signature in the genetic
variation within and among the affected lineages.
Thus, accurately estimating models of shared evolutionary history across
co-distributed taxa provides an important tool for better understanding how
regional and global biogeographic processes affect biodiversity.

@@

The software package \msb implements such an approach in an
approximate-Bayesian model-choice framework \citep{Hickerson2006,Huang2011}.
Given aligned sequence data for \npairs{} pairs of populations, \msb 
estimates the number of divergence events shared among the pairs, the timing of
the events, and the assignment of pairs to the events, while integrating out
uncertainty in demographic parameters and the genealogical history of the
sequences.
% samples over all the possible ways of assigning the divergences between the
% populations of each pair to 1--\npairs{} divergence events, while estimating
% the timing of the events and integrating out uncertainty in demographic
% parameters and the genealogical history of the sequences.
Thus, when the number of pairs is not trivial, the method samples over a large
number of divergence models of differing dimensionality (i.e., all the possible
partitions of \npairs{} pairs to 1--\npairs{} divergence-time parameters), and,
in so doing, estimates the posterior probability of each possible divergence
model.

For priors on divergence-time and demographic parameters, \msb uses continuous
uniform probability distributions.
This causes divergence models with more divergence-time parameters to
integrate over a \emph{much} greater parameter space with low likelihood yet
high prior density, resulting in small marginal likelihoods relative to models
with fewer divergence-time parameters \citep{Jeffreys1939,Lindley1957}.
Given that the marginal likelihood of a model weighted by its prior is what
determines its posterior probability, this causes \msb to support models with
fewer divergence events.
Furthermore, \msb uses a discrete uniform prior over the number of divergence
events $1,2,\ldots,\npairs{}$.
Because there are many more possible divergence models for intermediate numbers
of divergence events, this imposes a prior on divergence models that strongly
disfavors such models and puts most of the prior mass on models with either
very few or very many divergence-time parameters (see Figure 5 of
\citet{Oaks2012}; for brevity we will refer to this prior as ``U-shaped'').
Given that models with many divergence events have small marginal likelihoods
due to the uniform priors on divergence-time parameters, the U-shaped prior
will effectively create a strong prior preference for models with very few
divergence events.
Recently, \citet{Oaks2012} found via simulation that \msb will often strongly
support biogeographically interesting models with a small number of divergence
events shared among the taxa, even when divergences were random over broad
timescales.
They suggested this behavior was due to the combination of uniform priors on
parameters and the U-shaped prior on divergence models.

@@

We modify the \msb model to test whether alternative parameterizations and
priors will improve the behavior of the approximate-Bayesian model-choice
approach.
We implement a Dirichlet-process prior (DPP) over all possible models of
divergence, and use gamma and beta probability distributions in place of
uniform priors on many of the model's parameters.
Our simulation-based analyses show that the new implementation has improved
robustness, accuracy, and power compared to the original model.
Also, our results confirm that both the uniform priors on demographic and
divergence-time parameters and the U-shaped prior on divergence models
contribute to the biases of \msb.
After reanalyzing sequence data of 22 pairs of taxa from the Philippines
\citep{Oaks2012} under the new model, we find a large amount of posterior
uncertainty in the number of divergence events shared among the taxa; a result
congruent with intuition given the richness of the model and the relatively
small amount of information in the data.

@@

\subsubsection*{The priors on nuisance parameters}
Following the recommendations of \citet{Oaks2012}, we have replaced the use of
continuous uniform distributions for priors on many of the
model's parameters (\divTime{}, \ancestralTheta{}, \descendantTheta{1}{},
\descendantTheta{2}{}, \bottleScalar{1}{}, \bottleScalar{2}{},
\recombinationRate, \migrationRate{}) with more flexible parametric
distributions from the exponential family.
We introduce gamma-distributed priors for rate parameters that have a sample
space of all positive real numbers (\divTime{}, \ancestralTheta{},
\descendantTheta{1}{}, \descendantTheta{2}{}, \recombinationRate,
\migrationRate{}), and beta-distributed priors for parameters that are
proportions bounded by zero and one (\bottleScalar{1}{} and
\bottleScalar{2}{}).
These priors provide an investigator with much greater flexibility in
expressing prior uncertainty regarding the parameters of the model.

Another distinction between our model and that of \msb, is the prior on the
sizes of the descendant populations of each pair.
As described by \citet{Oaks2012}, \msb uses the joint prior
\begin{equation}
    \descendantTheta{1}{}, \descendantTheta{2}{} \sim
    Beta(1,1) \times 2 \times U(\uniformMin{\myTheta{}},
    \uniformMax{\descendantTheta{}{}}),
    \label{eq:jointThetaPrior}
\end{equation}
such that the user-specified uniform prior on descendant population
size is a prior on the \emph{mean} size of the two descendant
populations of each pair.
Under our model, the sizes of the descendant populations of each
pair are \iid as
$\descendantTheta{1}{} \sim Gamma(\gammaShape{\descendantTheta{}{}},
\gammaScale{\descendantTheta{}{}})$
and
$\descendantTheta{2}{} \sim Gamma(\gammaShape{\descendantTheta{}{}},
\gammaScale{\descendantTheta{}{}})$.
This relaxes the assumption that the sizes of the two descendant populations
are interdependent and negatively correlated.

>>>>

\section*{Introduction}
Understanding the processes that generate biodiversity and regulate
community assembly is a major goal of evolutionary biology.
Given the dynamic nature of our planet, a major component of this goal is
understanding how geological and climatic processes can affect the
evolutionary history of whole groups of co-distributed species and their
associated microbiota, including pathogens.
Such large-scale processes predict patterns of shared evolutionary history
across affected populations of organisms, e.g., temporally clustered
divergences across co-distributed species.
At recent timescales, such events should leave a signature in the genetic
variation within and among the affected lineages.
Thus, accurately estimating models of shared evolutionary history across
co-distributed taxa is important for better understanding how regional and
global biogeographic processes affect biodiversity.

@@

The software package \msb implements such an approach in an
approximate-Bayesian model-choice framework \citep{Hickerson2006,Huang2011}.
Given aligned sequence data for \npairs{} pairs of populations, \msb 
estimates the number of divergence events shared among the pairs, the timing of
the events, and the assignment of pairs to the events, while integrating out
uncertainty in demographic parameters and the genealogical history of the
sequences.
% samples over all the possible ways of assigning the divergences between the
% populations of each pair to 1--\npairs{} divergence events, while estimating
% the timing of the events and integrating out uncertainty in demographic
% parameters and the genealogical history of the sequences.
Thus, the method samples over all possible divergence models of differing
dimensionality (i.e., all the possible partitions of \npairs{} pairs to
1--\npairs{} divergence-time parameters), and, in so doing, estimates the
posterior probability of each model.

For priors on divergence-time and demographic parameters, \msb uses continuous
uniform probability distributions.
This causes divergence models with more divergence-time parameters to integrate
over a \emph{much} greater parameter space with low likelihood yet high prior
density, resulting in small marginal likelihoods relative to models with fewer
divergence-time parameters \citep{Jeffreys1939,Lindley1957}.
Given that the marginal likelihood of a model weighted by its prior is what
determines its posterior probability, this can cause support for models with
fewer divergence events \citep{Oaks2012}.
Furthermore, \msb uses a discrete uniform prior over the number of divergence
events $1,2,\ldots,\npairs{}$.
Because there are many more possible assignments of populations pairs to
intermediate numbers of divergence events, this imposes a prior on divergence
models that puts most of the prior mass on models with either very few or very
many divergence-time parameters (see Figure 5 of \citet{Oaks2012}; for brevity
we will refer to this prior as ``U-shaped'').
Given that models with many divergence events can have small marginal
likelihoods due to the uniform priors on divergence-time parameters, the
U-shaped prior will effectively create a strong prior preference for models
with very few divergence events.
Recently, \citet{Oaks2012} found via simulation that \msb will often strongly
support biogeographically interesting models with a small number of divergence
events shared among the taxa, even when divergences were random over broad
timescales.
They suggested this behavior was due to the combination of uniform priors on
parameters and the U-shaped prior on divergence models.

@@

We modify the \msb model to test whether alternative parameterizations and
priors will improve the behavior of the approximate-Bayesian model-choice
approach.
We implement a Dirichlet-process prior (DPP) over all possible models of
divergence, and use gamma and beta probability distributions in place of
uniform priors on many of the model's parameters.
Using simulations, we show that the new implementation has improved robustness,
accuracy, and power compared to the original model.
Also, our results confirm that the improved performance of the new model
is due to a combination of
(1) more flexible priors on divergence-time and demographic parameters that
avoid placing high prior density in improbable regions of parameter space, and
(2) a diffuse Dirichlet-process prior that does not strongly disfavor
divergence models with intermediate numbers of divergence events.
% Also, our results confirm that both the uniform priors on demographic and
% divergence-time parameters and the U-shaped prior on divergence models
% contribute to the tendency of \msb.
After reanalyzing sequence data from 22 pairs of taxa from the Philippines
\citep{Oaks2012} under the new model, we find a large amount of posterior
uncertainty in the number of divergence events shared among the taxa; a result
in contrast with the original \msb model \citep{Oaks2012} and congruent with
intuition given the richness of the model and the relatively small amount of
information in the data.

@@

\subsubsection*{The priors on nuisance parameters}
We have replaced the use of continuous uniform distributions for priors on many
of the model's parameters (\divTime{}, \ancestralTheta{},
\descendantTheta{1}{}, \descendantTheta{2}{}, \bottleScalar{1}{},
\bottleScalar{2}{}, \recombinationRate, \migrationRate{}) with more flexible
parametric distributions from the exponential family.
We introduce gamma-distributed priors for rate parameters that have a sample
space of all positive real numbers (\divTime{}, \ancestralTheta{},
\descendantTheta{1}{}, \descendantTheta{2}{}, \recombinationRate,
\migrationRate{}), and beta-distributed priors for parameters that are
proportions bounded by zero and one (\bottleScalar{1}{} and
\bottleScalar{2}{}).
These priors provide an investigator with much greater flexibility in
expressing prior uncertainty regarding the parameters of the model.

In addition, we have modified the prior on the sizes of the descendant
populations of each pair.
As described by \citet{Oaks2012}, \msb uses the joint prior
\begin{equation}
    \descendantTheta{1}{}, \descendantTheta{2}{} \sim
    Beta(1,1) \times 2 \times U(\uniformMin{\myTheta{}},
    \uniformMax{\descendantTheta{}{}}),
    \label{eq:jointThetaPrior}
\end{equation}
such that the user-specified uniform prior on descendant population
size is a prior on the \emph{mean} size of the two descendant
populations of each pair.
Under our model, the sizes of the descendant populations of each
pair are \iid as
$\descendantTheta{1}{} \sim Gamma(\gammaShape{\descendantTheta{}{}},
\gammaScale{\descendantTheta{}{}})$
and
$\descendantTheta{2}{} \sim Gamma(\gammaShape{\descendantTheta{}{}},
\gammaScale{\descendantTheta{}{}})$.
This relaxes the assumption that the sizes of the two descendant populations
are interdependent and negatively correlated.
